Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-04-21 Paper/manuscript_v2/manuscript_07_01.tex
Edit to sec 5.1

This example consists of a simulated dataset with $n=250$ ($i=1,2,\dots,n$) single draws from a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\bm{y}$, Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws.

The exact GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(\bm{0}, \sigma^2 \bm{I}) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
%
where $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function at the input values $x_i$, and $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, with $I$ representing the identity matrix. The previous formulation corresponds to the latent form of a GP model. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with the Mat{\'e}rn($\nu$=3/2) covariance function $k$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,2,\dots,n$.

@@

Figure~\ref{ch5_fig10_Posteriors_exI} shows the posteriors predictive distributions of the three models, the exact GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$; see equation (\ref{eq_boundary})), and the spline model with 80 knots. The true data-generative function and the noisy observations are also plotted. The sample observations are plotted as circles and the out-of-sample or test data, which have not been taking part on training the models, are plotted as crosses. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation. The posteriors of the three models, exact GP, HSGP and spline, are pretty similar within the interpolation input space. However, when extrapolating the spline model solution clearly differs from the exact GP and HSGP models as well as the actual observations.

@@

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{ch5_fig11_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{ch5_fig10_Posteriors_exI} and \ref{ch5_fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the exact GP model for interpolating and extrapolating data. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.

Figure~\ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The HSGP model is on average roughly 400 times faster than the exact GP and 10 times faster than the spline model, for this particular application with a univariate input space. Also, the increase in computation time as a function of the number of basis functions in a univariate input space is relatively slight.

@@

In the online supplemental material, some other case studies are presented. From those examples, it can be seen how computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Even though, in a bivariate input space, the computation time increases significantly with $D$, the HSGP model works significantly faster than the exact GP for most of the non-linear $2D$ functions (even highly wiggly functions; see Figures B.3-right and C.3 in the online material). However, HSGPs tend to be slower than exact GPs for $D>3$ with a relatively low number of basis functions ($m \gtrsim 5$), as well as even for $D=3$ with a moderate high number of basis functions ($m \gtrsim 20$; see Figure~C.3 in the online material). In all of the investigated cases, choosing the optimal boundary factor in the HSGP approximation reduces the number of required basis functions noticeably (see Figures~A.3, B.3-left and C.2 in the online material) and therefore also reduces computational time drastically in particular in multivariate input spaces.

>>>>

In this experiment, we analyze a synthetic dataset with $n = 250$ observations, where the true data generating process is a Gaussian process with additive noise. The data points are simulated from the model $y_i = f(x_i) + \epsilon_i$, where $f$ is a sample from a Gaussian process using the Mat{\'e}rn($\nu$=3/2) covariance function with marginal variance $\alpha=1$ and lengthscale $\ell=0.15$ at inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1]$. $\epsilon_i$ is additive Gaussian noise with standard deviation $\sigma=0.2$.

The exact GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(\bm{0}, \sigma^2 \bm{I}) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
%
where $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function at the input values $x_i$, and $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, with $\bm{I}$ representing the identity matrix. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with the Mat{\'e}rn($\nu$=3/2) covariance function $k$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,2,\dots,n$.

@@

Figure~\ref{ch5_fig10_Posteriors_exI} shows the posteriors predictive distributions of the three models, the exact GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$; see equation (\ref{eq_boundary})), and the spline model with 80 knots. The true data-generating function and the noisy observations are also plotted. The sample observations are plotted as circles and the out-of-sample or test data are plotted as crosses. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation. The posteriors of the three models, exact GP, HSGP and spline, are similar in the interpolation regions of the input space. However, when extrapolating the spline model solution clearly differs from the exact GP and HSGP models as well as the actual observations.

@@

In order to assess the performance of the models as a function of the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model, and different number of knots for the spline model, have been fitted. Figure \ref{ch5_fig11_MSE_exI_inter} shows the standardized root mean squared error (SRMSE) for interpolation and extrapolating data as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating model. From Figures \ref{ch5_fig10_Posteriors_exI} and \ref{ch5_fig11_MSE_exI_inter}, it is seen that the HSGP method yield a good approximation of the exact GP model for both interpolation and extrapolation. However, the spline model does not extrapolate data properly. Both models show roughly similar interpolating performance.

Figure~\ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The HSGP model is on average roughly 400 times faster than the exact GP and 10 times faster than the spline model for this particular model and data. Also, it is seen that the computation time increases slowly as a function of the number of basis functions.

@@

In the online supplemental material, additional case studies are presented. From those examples, it can be seen how the computation time of the HSGP model increases rapidly with the number of input dimensions ($D$) since the number of basis functions in the approximation increases exponentially with $D$ (see eq.~(\ref{eq_m_multi})). Even though, for a bivariate input space, the computation time increases significantly with $D$, the HSGP model works significantly faster than the exact GP for most of the non-linear $2D$ functions (even highly wiggly functions; see Figures B.3-right and C.3 in the online material). For moderate sized datasets, HSGPs tend to be slower than exact GPs for $D>3$ with a relatively low number of basis functions ($m \gtrsim 5$), as well as even for $D=3$ with a moderate high number of basis functions ($m \gtrsim 20$; see Figure~C.3 in the online material). However, the HSGP method will be computationally faster than the exact GP for larger datasets due the cubic scaling of the exact GP. In all of the investigated cases, choosing the optimal boundary factor in the HSGP approximation reduces the number of required basis functions noticeably (see Figures~A.3, B.3-left and C.2 in the online material) and therefore also reduces computational time drastically in particular in multivariate input spaces.
